DBES Grand Challenge : High throughput, scalable solution for processing taxi trip data

### 2015-03-19 paper/experiments.tex
updated the paper

Figure \ref{time_profit} shows the execution times for both setups. In both setups, there is a higher latency at one instance due to lack of parallelism. Then for distributed set up execution times get reduced with addition of new resources and for single JVM setup it remains a constant. This shows our solution scales up in a distributed setup. We can increase the system throughput by adding new nodes until \textit{ProfitEventEmitter} becomes the bottleneck.

@@

Figure \ref{latency_profit} shows the latencies. For single JVM, latencies slightly increased with the addition of new instances as expected. But for distributed setups there is a higher latencies compared to others. This latency is due to longer queues at the \textit{TopProfitProcessor}. In our message ordering process, we can only send a message to \textit{TopProfitProcessor} only if all queues are not empty. This allows some processes with less top ten profitable events to delay other process messages. This problem can  possibly fixed by adding a limit to the message queue size. On the other hand in a real time system, this event ordering many not required since event processing happens at real time.

>>>>

Figure \ref{time_profit} shows the execution times for both setups. In both setups, there is a higher execution time at one instance due to lack of parallelism. Then for distributed set up execution times get reduced with addition of new resources and for single JVM setup it remains a constant. This shows our solution scales up in a distributed setup. We can increase the system throughput by adding new nodes until \textit{ProfitEventEmitter} becomes the bottleneck.

@@

Figure \ref{latency_profit} shows the latencies. For single JVM, latencies slightly increased with the addition of new instances as expected. But for distributed setups there is a higher latencies compared to others. This latency is due to longer queues at the \textit{TopProfitProcessor}. In our message ordering process, we can send a message to \textit{TopProfitProcessor} only if all queues are not empty. This allows some processes with less top ten profitable events to delay other process messages. This problem can  possibly fixed by adding a limit to the message queue size. On the other hand in a real time system, this event ordering many not required since event processing happens at real time.

### 2015-03-19 paper/query_evaluation.tex
updated the paper

We can sequentially execute events either in a single Java Virtual Machine (JVM) or different JVMs (either in one machine or different machines) using TCP sockets to communicate between process instances. Figure \ref{sequential} shows such possible executions.

@@

\begin{enumerate}
	\item \textbf{Expired values} : Route count of a particular route can be reduced due to expiration of some events. For an example, lets assume route count for route R1 is 10 and R1 is included in a top ten route event generated by a \textit{QueryEvaluator}. Then lets assume route count for R1 has been reduced to 7 (due to expiration events), when the next event is generated and hence R1 is not included in the next top ten route event. However since R1 value still is in the \textit{ResultAggregator} that may produce an event with this value. To avoid this problem, we send the removed route set details with the top ten route event generated by the \textit{QueryEvaluator}. In that case \textit{ResultAggregator} can first remove those routes.
	\item \textbf{Out of order events} : \textit{ResultAggregator} may receive events in a different order than those are produced at the \textit{EventEmitter}. This out of order events may change the final output of the \textit{ResultAggregator}. We order messages at the \textit{ResultAggregator} to avoid this problem. In order to order messages, we keep a message queue for each \textit{QueryEvaluator} and sent the message with least sequence number to \textit{ResultAggregator},  if a message is present in all queues. 
	\item \textbf{Intermediate top ten route events} : A route event increases the route count for that events' route. Further that can generate some expiration events which may reduce the route count for other routes. The top ten frequent routes can be changed due to these expiration events as well. Lets say in one \textit{QuerayEvaluator} case, a particular event cause two routes R1 and R2 to change and generates a top ten route change event. In the distributed setup these two routes can be in two \textit{QueryEvaluators} and that can cause two top ten route change events. These two events may result in two top ten route change events at \textit{ResultAggregator} as well instead of one. Furether we observe only around 5\% of such events generated, and system goes to the one \textit{QueryEveluator} result after all such event are being executed. Solution for such a problem may depend on the nature in which a real application, uses these results in decision making process. Therefore we could not provide any fix for this problem.
\end{enumerate}

>>>>

We can sequentially execute events either in a single Java Virtual Machine (JVM) or different JVMs (either in one machine or different machines) using TCP sockets to communicate among process instances. Figure \ref{sequential} shows such possible executions.

@@

\begin{enumerate}
	\item \textbf{Expired values} : Route count of a particular route can be reduced due to expiration of some events. For an example, lets assume route count for route R1 is 10 and R1 is included in a top ten route event generated by a \textit{QueryEvaluator}. Then lets assume route count for R1 has been reduced to 7 (due to expiration events), when the next event is generated and hence R1 is not included in the next top ten route event. However since R1 value still is in the \textit{ResultAggregator} that may produce an event with this value. To avoid this problem, we send the removed route set details with the top ten route event generated by the \textit{QueryEvaluator}. In that case \textit{ResultAggregator} can first remove those routes.
	\item \textbf{Out of order events} : \textit{ResultAggregator} may receive events in a different order than those are produced at the \textit{EventEmitter}. This out of order events may change the final output of the \textit{ResultAggregator}. We order messages at the \textit{ResultAggregator} to avoid this problem. In order to order messages, we keep a message queue for each \textit{QueryEvaluator} and sent the message with least sequence number to \textit{ResultAggregator},  if a message is present in all queues. 
	\item \textbf{Intermediate top ten route events} : A route event increases the route count for that events' route. Further that can generate some expiration events which may reduce the route count for other routes. The top ten frequent routes can be changed due to these expiration events as well. Lets say in one \textit{QuerayEvaluator} case, a particular event cause two routes R1 and R2 to change and generates a top ten route change event. In the distributed setup these two routes can be in two \textit{QueryEvaluators} and that can cause two top ten route change events. These two events may result in two top ten route change events at \textit{ResultAggregator} as well instead of one. Further we observed only around 5\% of such events are generated, and system goes to the one \textit{QueryEveluator} result after all such event are being executed. Solution for such a problem may depend on the nature in which a real application, uses these results in decision making process. Therefore we could not provide any fix for this problem.
\end{enumerate}

### 2015-03-19 paper/query_implementation.tex
updated the paper

One way of implementing this data structure is to use a heap. Heap supports adding and extracting maximum value operations in O(log(n)) time. However since this application requires top 10 values, we need to retrieve maximum value ten times and insert them back. On the other hand if we use a doubly linked list to keep values, it will take O(n) time for  \texttt{decrementPostion} and \texttt{incrementPosition} operations. Therefore we can use a doubly linked list to keep first 10 values (this makes time complexity for all operations with this data structure O(1)) and a heap to keep other values to avoid above problems.  A Map which points to values can be used to retrieve values using a key.

@@

Table \ref{orderedlist_api} shows the operations of the \textit{OrderedList} data structure. First six methods has the same meanings as the \textit{NodeList} data structure and last two can be used to get the size of the list and the value of the last object. We internally keep a counter and a pointer to tail to support these operations in O(1) time. Figure \ref{orderedlist_impl} shows the implementation details. The Map structure (an HashMap) is used to retrieve object pointers with O(1) expected time. Each node of the doubly linked list points to previous and next nodes while storing the \textit{NodeValue}.

@@

Table \ref{dynamicheap_api} shows the operations of the \textit{DynamicHeap} data structure. \texttt{MoveUp} and \texttt{moveDown} operations are used to move an element up in the heap if the value increased, or move down in the heap if the value decreased. Figure \ref{dynamicheap_impl} shows the implementation details. Conventional heaps only support \texttt{add} and \texttt{extractMax} operations since heap operations requires to know the element position in heap. When adding an element we add it to last position and move that up and extracting maximum element start at array position 0. In order to support \texttt{moveUp}, \texttt{moveDown} and \texttt{remove} operations, we keep the array index with the object itself. This allows us to retrieve array index in O(1) time and move the element O(log(n)) time.

@@

\begin{algorithm}
\caption{Algorithm to generate top 10 frequent route change events}
\label{frequent_route_algorithm}
\begin{algorithmic} 
\STATE beforeTopTen $\leftarrow$ nodeList.getTopValues() 
\STATE window.add(event) 
\IF{ nodeList.containsKey(event.route) }
	\STATE routCount $\leftarrow$ nodeList.get(event.route) 
	\STATE routCount.count++ 
	\STATE nodeList.decrementPosition(event.route) 
\ELSE
	\STATE nodeList.add(event.route, new RouteCount(1)) 
\ENDIF

@@

>>>>

One way of implementing this data structure is to use a heap. Heap supports adding and extracting maximum value operations in O(log(n)) time. However since this application requires top 10 values, we need to retrieve maximum value ten times and insert them back. On the other hand if we use a doubly linked list to keep values, it will take O(n) time for  \texttt{decrementPostion} and \texttt{incrementPosition} operations. Therefore we can use a doubly linked list to keep first 10 values (this makes time complexity for all operations with this data structure O(1)) and a heap to keep other values to avoid above problems.  A \textit{Map} which points to values can be used to retrieve values using a key.

@@

Table \ref{orderedlist_api} shows the operations of the \textit{OrderedList} data structure. First six methods has the same meanings as the \textit{NodeList} data structure and last two can be used to get the size of the list and the value of the last object. We internally keep a counter and a pointer to tail to support these operations in O(1) time. Figure \ref{orderedlist_impl} shows the implementation details. The \textit{Map} structure (an \textit{java.util.HashMap}) is used to retrieve object pointers with O(1) expected time. Each node of the doubly linked list points to previous and next nodes while storing the \textit{NodeValue}.

@@

Table \ref{dynamicheap_api} shows the operations of the \textit{DynamicHeap} data structure. \texttt{MoveUp} and \texttt{moveDown} operations are used to move an element up in the heap if the value increased, or move down in the heap if the value decreased. Figure \ref{dynamicheap_impl} shows the implementation details. Conventional heaps only support \texttt{add} and \texttt{extractMax} operations since heap operations requires to know the element position in heap. When adding an element, we add it to last position and move that up and extracting maximum element start at array position 0. In order to support \texttt{moveUp}, \texttt{moveDown} and \texttt{remove} operations, we keep the array index with the object itself. This allows us to retrieve array index in O(1) time and move the element O(log(n)) time.

@@

\begin{algorithm}
\caption{Algorithm to generate top 10 frequent route change events}
\label{frequent_route_algorithm}
\begin{algorithmic}
\small 
\STATE beforeTopTen $\leftarrow$ nodeList.getTopValues() 
\STATE window.add(event) 
\IF{ nodeList.containsKey(event.route) }
	\STATE routCount $\leftarrow$ nodeList.get(event.route) 
	\STATE routCount.count++ 
	\STATE nodeList.decrementPosition(event.route) 
\ELSE
	\STATE nodeList.add(event.route, new RouteCount(1)) 
\ENDIF

@@

Algorithm \ref{pickup_algorithm} shows the logic to process pickup data. When a taxi gets a pickup, the taxi fare is added to that cell. We use \textit{ProfitCellNode} to keep the fare and empty taxi details to calculate the profitability. Then if its' taxi dropoff event has not been occurred (i.e this taxi has come to this location less than 30 minutes ago) it removes the taxi from that location. When the event get expired, it removes the fare from the \textit{ProfitCellNode}.

\begin{algorithm}
\caption{Algorithm to process pickup data}
\label{pickup_algorithm}
\begin{algorithmic}
\small
\STATE beforeTopTen $ \leftarrow $ nodeList.getTopValues()
\STATE paymentWindow.add(event)
\IF {nodeList.containsKey(event.pickUpCell)}
	\STATE profitCellNode $ \leftarrow $ nodeList.get(event.pickUpCell)
	\STATE profitCellNode.addFare(event.fare)
	\IF {profitCellNode.containsTaxi(event.medalliion)}
		\STATE profitCellNode.emptyTaxi--
		\STATE profitCellNode.removeTaxi(event.medallion)
	\ENDIF
	\IF {profitablity has increased}
		\STATE nodeList.decrementPosition(event.pickUpCell)
	\ELSE
		\STATE nodeList.incrementPosition(event.pickUpCell)	
	\ENDIF
\ELSE
	\STATE nodeList.add(event.pickUpCell, new ProfitCellNode(event.fare))
\ENDIF
\WHILE {there exits expired events}
	\STATE expiredEvent $ \leftarrow $ paymentWindow.poll()
	\STATE profitCellNode $ \leftarrow $ nodeList.get(expiredEvent.pickUpCell)
	\STATE profitCellNode.removeFare(expiredEvent.fare)

	\IF {profitCellNode is empty}
		\STATE nodeList.remove(expiredEvent.pickUpCell)
	\ELSIF {profitablity increased}
		\STATE nodeList.decrementPosition(event.pickUpCell)
	\ELSE
		\STATE nodeList.incrementPosition(event.pickUpCell)
	\ENDIF
\ENDWHILE
\STATE nowTopTen $ \leftarrow $ nodeList.getTopValues()
\IF {beforeTopTen not equals to nowTopTen}
	\STATE emit a top ten change event
\ENDIF

Algorithm \ref{dropoff_algorithm} shows the logic to process dropoff data. When a taxi trip ended in a location, number of empty taxis of that location has to be increased. If the taxi does not get a pickup within 30 mintes, then that taxi has to be removed from that location.

\begin{algorithm}
\caption{Algorithm to process dropoff data}
\label{dropoff_algorithm}
\begin{algorithmic}
\small
\STATE beforeTopTen $ \leftarrow $ nodeList.getTopValues()
\STATE dropWindow.add(event)
\IF {nodeList.containsKey(event.dropOffCell)}
	\STATE profitCellNode $ \leftarrow $ nodeList.get(event.dropOffCell)
	\STATE profitCellNode.addTaxi(event.medallion)
	\STATE profitCellNode.emptyTaxi++
	\STATE nodeList.incrementPosition(event.dropOffCell)

\ELSE
	\STATE nodeList.add(event.dropOffCell, new ProfitCellNode(event.medallion))
\ENDIF
\WHILE {there are expired events}
	\STATE expiredEvent $ \leftarrow $ dropWindow.poll()
	\STATE profitCellNode $ \leftarrow $ nodeList.get(expiredEvent.dropOffCell)
	\IF {profitCellNode.containsTaxi(expiredEvent.medallion)}
		\STATE profitCellNode.emptyTaxi--
		\STATE profitCellNode.removeTaxi(expiredEvent.medallion)

		\IF {profitCellNode is empty}
			\STATE nodeList.remove(event.dropoffCell)
		\ELSE 
			\STATE nodeList.decrementPosition(event.dropOffCell) 
		\ENDIF
	\ENDIF
\ENDWHILE
